Minimal trap spaces of Boolean networks are maximal siphons of their Petri net encoding

### 2022-12-22 manuscripts/TCS2022/tcs2022.tex
Result tables

As a computational application, by Theorem~\ref{theo:min_ts_2_max_sp}, we can reduce the problem of computing all minimal trap spaces of a Boolean network to the problem of computing all maximal conflict-free siphons of its Petri net encoding.
Note that in the case of stable states, this can be put in regard to the classical relationship between siphons and deadlocks in Petri nets.
It might actually be possible to generalize our result to any 1-safe place-complementary Petri net to define a notion of trap spaces that might be useful for the analysis of Petri nets, but this is out of the scope of this article.

It is noted that there are no existing methods specifically designed for computing maximal conflict-free siphons (even maximal siphons) of a Petri net. The reason might be that researchers mainly focus on minimal siphons~\cite{DBLP:journals/isci/LiuB16}. Hence, we here propose a new method based on Answer Set Programming (ASP)~\cite{DBLP:journals/aicom/GebserKKOSS11} for computing maximal conflict-free siphons of a Petri net. The details of the proposed method shall be given in the next section.

@@

\subsection{Constraint Satisfaction Problem}
\label{subsec:computation_csp}

The following Boolean CSP directly derives from the above characterization:
\begin{definition}

  Given a Petri net \(\mathcal{P} = (P, T, W)\) encoding a Boolean model \(\mathcal{M} = (V, F)\).
  The CSP \(\mathcal{C}(\mathcal{P})\) is the triple \((R, D, C)\) where
  \begin{itemize}%
    \item \(R = P\), i.e., a variable is introduced for each place of \(\mathcal{P}\),
    \item \(D(p) = \mathbb{B}\) for all \(p \in R\), i.e., the variables are Boolean,
    \item \(C = \{\neg p_v \vee \neg \overline{p_v} = 1 \mid \forall v \in V\} \bigwedge
\{(p = 1 \rightarrow \bigvee_{p' \in pred(t)}p' = 1) \mid p \in P, t \in pred(p)\}\).
  \end{itemize}

@@

Another possible method is to translate the characterization shown in Subsection~\ref{subsec:siphon_characterization} into the ASP \(\mathcal{L}\) as follows.
We introduce atom \verb|p-v| (resp. \verb|n-v|) to denote place \(p_v\) (resp. \(\overline{p_v}\)), \(\forall v \in V\). 
The set of all atoms in \(\mathcal{L}\) is given as \(\mathcal{A} = \bigcup_{v \in V}\{\verb|p-v|, \verb|n-v|\}\).
For each pair \((p, t)\) where \(p \in P, t \in T, t \in pred(p)\), we translate the rule (\ref{eq:siphon}) into the ASP rule
\[
\verb|a_1; ... ; a_k :- a.|
\]
where \(\verb|a| \in \mathcal{A}\) is the atom representing place \(p\) and \(\{\verb|a_1|, \dots,\verb|a_k|\} \subseteq \mathcal{A}\) is the set of atoms representing places in \(pred(t)\). The rule (\ref{eq:conflict}) is translated into the ASP rule
\[\verb|:- p-v, n-v.|\]
for each \(v \in V\). 
This ASP rule guarantees that two places representing the same node in \(\mathcal{M}\) never belong to the same siphon of \(\mathcal{P}\), representing the conflict-freeness.
Naturally, a Herbrand model (see, e.g,~\cite{DBLP:journals/aicom/GebserKKOSS11}) of \(\mathcal{L}\) is equivalent to a conflict-free siphon of \(\mathcal{P}\).
To guarantee that a Herbrand model is also a stable model (an answer set), we need to add to \(\mathcal{L}\) the two choice rules
\[
\verb|{p-v}. {n-v}.|
\]
for each \(v \in V\).
Note that the number of atoms of \(\mathcal{L}\) is only \(2n\), whereas the ASP encoding shown in~\cite{klarner2015computing} has as many atoms as the number of prime-implicants of the Boolean network and that number might be exponential in \(n\).
In~\cite{DBLP:conf/ictai/ChevalierFPZ19}, there is an ASP characterization of trap spaces that does not rely on minimal DNFs either and thus seems very similar to our ASP encoding.
Remarkably it only requires the DNF for the \emph{activation} part, using the information that it will only be used for locally-monotonic Boolean networks.
We would therefore expect that, when available, it will have comparable performance on the ASP part (the ASP program would be approximately twice smaller, though redundancy is not always bad in that field), but can also avoid combinatorial explosion of the Petri net encoding for some formula where the activation DNF is simple but the inhibition is not.
Since mpbn is included in our benchmark this will be evaluated in our experiments.

@@

@@

To assess the efficiency of the proposed method, implemented as a Python package named Trappist, we compare it with the state-of-the-art method implemented in the tool PyBoolNet~\cite{klarner2015computing,klarner2017pyboolnet} on both its own repository of models and large models from the literature.
In addition, we also include the tool mpbn~\cite{Paulev2020} to the benchmarks although it only handles locally-monotonic models, whereas both Trappist and PyBoolNet can handle general models.

@@

\setlength{\tabcolsep}{4pt}
\begin{table}[!htb]
  \caption{Timing comparisons between PyBoolNet, mpbn and Trappist on the PyBoolNet repository.
    Column \(n\) denotes the number of nodes of each model. Column \(|M|\) denotes the number of minimal trap spaces and for each method is given the computation time in seconds, asking only for the first 1000 trap spaces.
    A number in bold indicates a ratio greater than three compared to the best result. ``NM'' indicates a non-locally-monotonic model.}
  \centering%
  \label{tab:pyboolnet_repo}
  \begin{tabular}{rlrrrrrrr}
    \toprule
    & model & $n$ & $|M|$ & PyBoolNet & mpbn & Trappist \\ \midrule
    \rownb & arellano\_rootstem & 9 & 4 & 0.05 & 0.01 & \textbf{0.20}\\
    \rownb & calzone\_cellfate & 28 & 27 & 0.03 & \textbf{NM} & 0.03\\
    \rownb & dahlhaus\_neuroplastoma & 23 & 32 & 0.05 & 0.02 & 0.03\\
    \rownb & davidich\_yeast & 10 & 12 & 0.04 & 0.01 & 0.02\\
    \rownb & dinwoodie\_life & 15 & 7 & 0.03 & 0.01 & 0.01\\
    \rownb & dinwoodie\_stomatal & 13 & 1 & 0.03 & 0.01 & 0.01\\
    \rownb & faure\_cellcycle & 10 & 2 & 0.04 & 0.01 & 0.02\\
    \rownb & grieco\_mapk & 53 & 18 & 0.04 & 0.02 & 0.02\\
    \rownb & irons\_yeast & 18 & 1 & 0.05 & 0.01 & 0.02\\
    \rownb & jaoude\_thdiff & 103 & $>1000$ & \textbf{1.44} & \textbf{0.90} & 0.09\\
    \rownb & klamt\_tcr & 40 & 8 & 0.04 & 0.01 & 0.03\\
    \rownb & krumsiek\_myeloid & 11 & 6 & 0.03 & 0.01 & 0.01\\
    \rownb & multivalued & 13 & 4 & 0.03 & 0.01 & 0.01\\
    \rownb & n12c5 & 11 & 5 & \textbf{35.16} & 0.01 & 0.02\\
    \rownb & n3s1c1a & 2 & 2 & 0.02 & 0.01 & 0.01\\
    \rownb & n3s1c1b & 2 & 2 & 0.02 & 0.01 & 0.01\\
    \rownb & n5s3 & 4 & 3 & 0.03 & \textbf{NM} & 0.01\\
    \rownb & n6s1c2 & 5 & 3 & 0.03 & 0.01 & 0.01\\
    \rownb & n7s3 & 6 & 3 & 0.02 & 0.01 & 0.01\\
    \rownb & raf & 3 & 2 & 0.02 & 0.01 & 0.01\\
    \rownb & randomnet\_n15k3 & 15 & 3 & 0.03 & \textbf{NM} & 0.01\\
    \rownb & randomnet\_n7k3 & 7 & 10 & 0.03 & \textbf{NM} & 0.01\\
    \rownb & remy\_tumorigenesis & 34 & 25 & \textbf{2.14} & 0.02 & 0.02\\
    \rownb & saadatpour\_guardcell & 13 & 1 & 0.03 & 0.01 & 0.01\\
    \rownb & selvaggio\_emt & 56 & $>1000$ & \textbf{1.02} & \textbf{0.52} & 0.09\\
    \rownb & tournier\_apoptosis & 12 & 3 & 0.04 & 0.01 & 0.01\\
    \rownb & xiao\_wnt5a & 7 & 4 & 0.03 & 0.01 & 0.01\\
    \rownb & zhang\_tlgl & 60 & 156 & 0.22 & \textbf{NM} & 0.05\\
    \rownb & zhang\_tlgl\_v2 & 60 & 258 & 0.09 & 0.15 & 0.02\\
    \bottomrule
  \end{tabular}
\end{table}
\setcounter{rownumber}{0}

@@

@@

>>>>

As a computational application, by Theorem~\ref{theo:min_ts_2_max_sp}, we can reduce the problem of computing all minimal trap spaces of a Boolean network to the problem of computing all maximal conflict-free siphons of its Petri net encoding.
Note that in the case of special types of trap spaces (e.g., fixed points), this can be put in regard to special types of siphons in Petri nets.
See Subsection~\ref{subsec:computation_special_ts} for more discussions about many special types of trap spaces.
It might actually be possible to generalize our result to any 1-safe place-complementary Petri net to define a notion of trap spaces that might be useful for the analysis of Petri nets, but this is out of the scope of this article.

It is noted that there are no existing methods specifically designed for computing maximal conflict-free siphons (even maximal siphons) of a Petri net.
The reason might be that researchers mainly focus on minimal generic siphons~\cite{DBLP:journals/isci/LiuB16} in the field of Petri nets. 
Hence, we here propose several methods for computing maximal conflict-free siphons of a Petri net. 
The details of the proposed methods shall be given in the next section.

@@

\subsection{Constraint satisfaction problem}
\label{subsec:computation_csp}

The following Boolean Constraint Satisfaction Problem (CSP) directly derives from the above characterization:
\begin{definition}

  Given a Petri net \(\mathcal{P} = (P, T, W)\) encoding a Boolean network \(\mathcal{N} = (V, F)\).
  The CSP \(\mathcal{C}(\mathcal{P})\) is the triple \((R, D, C)\) where
  \begin{itemize}%
    \item \(R = P\), i.e., a variable is introduced for each place of \(\mathcal{P}\),
    \item \(D(p) = \mathbb{B}\) for all \(p \in R\), i.e., the variables are Boolean,
    \item \(C = \{\neg p_v \vee \neg \overline{p_v} = 1 \mid \forall v \in V\} \bigwedge
\{(p = 1 \rightarrow \bigvee_{p' \in pred(t)}p' = 1) \mid p \in P, t \in pred(p)\}\).
  \end{itemize}

@@

Another possible method is to translate the characterization shown in Subsection~\ref{subsec:siphon_characterization} into the ASP \(\mathcal{L}\) as follows.
We introduce atom \verb|p-v| (resp. \verb|n-v|) to denote place \(p_v\) (resp. \(\overline{p_v}\)), \(\forall v \in V\). 
The set of all atoms in \(\mathcal{L}\) is given as \(\mathcal{A} = \bigcup_{v \in V}\{\verb|p-v|, \verb|n-v|\}\).
For each pair \((p, t)\) where \(p \in P, t \in T, t \in pred(p)\), we translate the rule (\ref{eq:siphon}) into the ASP rule
\[
\verb|a_1; ... ; a_k :- a.|
\]
where \(\verb|a| \in \mathcal{A}\) is the atom representing place \(p\) and \(\{\verb|a_1|, \dots,\verb|a_k|\} \subseteq \mathcal{A}\) is the set of atoms representing places in \(pred(t)\). The rule (\ref{eq:conflict}) is translated into the ASP rule
\[\verb|:- p-v, n-v.|\]
for each \(v \in V\). 
This ASP rule guarantees that two places representing the same node in \(\mathcal{M}\) never belong to the same siphon of \(\mathcal{P}\), representing the conflict-freeness.
Naturally, a Herbrand model (see, e.g.,~\cite{DBLP:journals/aicom/GebserKKOSS11}) of \(\mathcal{L}\) is equivalent to a conflict-free siphon of \(\mathcal{P}\).
To guarantee that a Herbrand model is also a stable model (an answer set), we need to add to \(\mathcal{L}\) the two choice rules
\[
\verb|{p-v}. {n-v}.|
\]
for each \(v \in V\).
Note that the number of atoms of \(\mathcal{L}\) is only \(2n\), whereas the ASP encoding shown in~\cite{klarner2015computing} has as many atoms as the number of prime-implicants of the Boolean network and that number might be exponential in \(n\).
In~\cite{DBLP:conf/ictai/ChevalierFPZ19}, there is an ASP characterization of trap spaces that does not rely on minimal DNFs either and thus seems very similar to our ASP encoding.
Remarkably it only requires the DNF for the \emph{activation} part, using the information that it will only be used for locally-monotonic Boolean networks.
We would therefore expect that, when available, it will have comparable performance on the ASP part (the ASP program would be approximately twice smaller, though redundancy is not always bad in that field), but can also avoid combinatorial explosion of the Petri net encoding for some formula where the activation DNF is simple but the inhibition is not.
Since mpbn is included in our benchmark this will be evaluated in our experiments.

@@

\subsection{Computation of special types of trap spaces}
\label{subsec:computation_special_ts}

@@

To assess the efficiency of the proposed method, implemented as a Python package named Trappist, we compared it with the state-of-the-art method implemented in the tool PyBoolNet~\cite{klarner2015computing,klarner2017pyboolnet} on both its own repository of models and large models from the literature.
In addition, we also included the tool mpbn~\cite{Paulev2020} to the benchmarks although it only handles locally-monotonic models, whereas both Trappist and PyBoolNet can handle general models.
To obtain a more comprehensive comparison, we also used random models generated by a third-party software (i.e., BoolNet R package~\cite{mussel2010boolnet}).

@@

\setlength{\tabcolsep}{4pt}
\begin{table}[!htb]
  \caption{Timing comparisons between PyBoolNet, mpbn and Trappist on the PyBoolNet repository.
    Column \(n\) denotes the number of nodes of each model. Column \(|M|\) denotes the number of minimal trap spaces and for each method is given the computation time in seconds, asking only for the first 1000 trap spaces.
    A number in bold indicates a ratio greater than three compared to the best result. ``NM'' indicates a non-locally-monotonic model.}
  \centering%
  \label{tab:pyboolnet_repo}
  \begin{tabular}{rlrrrrrr}
    \toprule
    & model & $n$ & $|M|$ & LQM & PBN & mpbn & Trappist \\ \midrule
    \rownb & arellano\_rootstem & 9 & 4 & - & 0.05 & 0.01 & \textbf{0.20}\\
    \rownb & calzone\_cellfate & 28 & 27 & - & 0.03 & \textbf{NM} & 0.03\\
    \rownb & dahlhaus\_neuroplastoma & 23 & 32 & - & 0.05 & 0.02 & 0.03\\
    \rownb & davidich\_yeast & 10 & 12 & - & 0.04 & 0.01 & 0.02\\
    \rownb & dinwoodie\_life & 15 & 7 & - & 0.03 & 0.01 & 0.01\\
    \rownb & dinwoodie\_stomatal & 13 & 1 & - & 0.03 & 0.01 & 0.01\\
    \rownb & faure\_cellcycle & 10 & 2 & - & 0.04 & 0.01 & 0.02\\
    \rownb & grieco\_mapk & 53 & 18 & - & 0.04 & 0.02 & 0.02\\
    \rownb & irons\_yeast & 18 & 1 & - & 0.05 & 0.01 & 0.02\\
    \rownb & jaoude\_thdiff & 103 & $>1000$ & - & \textbf{1.44} & \textbf{0.90} & 0.09\\
    \rownb & klamt\_tcr & 40 & 8 & - & 0.04 & 0.01 & 0.03\\
    \rownb & krumsiek\_myeloid & 11 & 6 & - & 0.03 & 0.01 & 0.01\\
    \rownb & multivalued & 13 & 4 & - & 0.03 & 0.01 & 0.01\\
    \rownb & n12c5 & 11 & 5 & - & \textbf{35.16} & 0.01 & 0.02\\
    \rownb & n3s1c1a & 2 & 2 & - & 0.02 & 0.01 & 0.01\\
    \rownb & n3s1c1b & 2 & 2 & - & 0.02 & 0.01 & 0.01\\
    \rownb & n5s3 & 4 & 3 & - & 0.03 & \textbf{NM} & 0.01\\
    \rownb & n6s1c2 & 5 & 3 & - & 0.03 & 0.01 & 0.01\\
    \rownb & n7s3 & 6 & 3 & - & 0.02 & 0.01 & 0.01\\
    \rownb & raf & 3 & 2 & - & 0.02 & 0.01 & 0.01\\
    \rownb & randomnet\_n15k3 & 15 & 3 & - & 0.03 & \textbf{NM} & 0.01\\
    \rownb & randomnet\_n7k3 & 7 & 10 & - & 0.03 & \textbf{NM} & 0.01\\
    \rownb & remy\_tumorigenesis & 34 & 25 & - & \textbf{2.14} & 0.02 & 0.02\\
    \rownb & saadatpour\_guardcell & 13 & 1 & - & 0.03 & 0.01 & 0.01\\
    \rownb & selvaggio\_emt & 56 & $>1000$ & - & \textbf{1.02} & \textbf{0.52} & 0.09\\
    \rownb & tournier\_apoptosis & 12 & 3 & - & 0.04 & 0.01 & 0.01\\
    \rownb & xiao\_wnt5a & 7 & 4 & - & 0.03 & 0.01 & 0.01\\
    \rownb & zhang\_tlgl & 60 & 156 & - & 0.22 & \textbf{NM} & 0.05\\
    \rownb & zhang\_tlgl\_v2 & 60 & 258 & - & 0.09 & 0.15 & 0.02\\
    \bottomrule
  \end{tabular}
\end{table}
\setcounter{rownumber}{0}

@@

\subsection{BBM repository}

\setlength{\tabcolsep}{4pt}
\begin{table}[!h]
  \caption{Results on the real-world models from the BBM repository.}
  \centering%
  \label{tab:bbm_repo}
  \begin{tabular}{l||r|r|r||r}
    \toprule
    Method & 1 min & 2 min & 3 min & Time average (s) \\ \midrule
    PyBoolNet & - & - & - & 0.05 \\
    mpbn & - & - & - & 0.05 \\
    Trappist & - & - & - & 0.05 \\
    \bottomrule
  \end{tabular}
\end{table}
\setcounter{rownumber}{0}

@@

\begin{table}[!htb]
  \caption{Results on N-K models.}
  \centering%
  \label{tab:result_N_K}
  \begin{tabular}{rrrrrrr}
    \toprule
    & \multicolumn{2}{c}{1 min} & \multicolumn{2}{c}{2 min} & \multicolumn{2}{c}{3 min} \\
    \cmidrule(lr){2-7}
    \(n\) & PBN & Trappist & PBN & Trappist & PBN & Trappist \\
    \midrule
    100 & - & - & - & - & - & - \\
    150 & - & - & - & - & - & - \\
    200 & - & - & - & - & - & - \\
    250 & - & - & - & - & - & - \\
    300 & - & - & - & - & - & - \\
    350 & - & - & - & - & - & - \\
    400 & - & - & - & - & - & - \\
    \bottomrule
  \end{tabular}
\end{table}

\begin{table}[!htb]
  \caption{Results on scale-free models.}
  \centering%
  \label{tab:result_N_K}
  \begin{tabular}{rrrrrrr}
    \toprule
    & \multicolumn{2}{c}{1 min} & \multicolumn{2}{c}{2 min} & \multicolumn{2}{c}{3 min} \\
    \cmidrule(lr){2-7}
    \(n\) & PBN & Trappist & PBN & Trappist & PBN & Trappist \\
    \midrule
    500 & - & - & - & - & - & - \\
    750 & - & - & - & - & - & - \\
    1000 & - & - & - & - & - & - \\
    1250 & - & - & - & - & - & - \\
    1500 & - & - & - & - & - & - \\
    1750 & - & - & - & - & - & - \\
    2000 & - & - & - & - & - & - \\
    \bottomrule
  \end{tabular}
\end{table}
